---
title: "15 Insights from Colorado School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from Colorado School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  eval = TRUE,
  cache = TRUE
)
```

```{r load-packages}
library(coschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))

# Load pre-computed data bundled with the package.
# This ensures vignettes build reliably in CI without network access.
# Falls back to live fetch if bundled data is unavailable.
enr <- tryCatch(
  readRDS(system.file("extdata", "enrollment_bundled.rds", package = "coschooldata")),
  error = function(e) {
    warning("Bundled data not found, fetching live data")
    fetch_enr_multi(c(2020, 2024), use_cache = TRUE)
  }
)
if (is.null(enr) || nrow(enr) == 0) {
  enr <- fetch_enr_multi(c(2020, 2024), use_cache = TRUE)
}

enr_2024 <- tryCatch(
  readRDS(system.file("extdata", "enrollment_2024_tidy.rds", package = "coschooldata")),
  error = function(e) {
    warning("Bundled data not found, fetching live data")
    fetch_enr(2024, use_cache = TRUE)
  }
)
if (is.null(enr_2024) || nrow(enr_2024) == 0) {
  enr_2024 <- fetch_enr(2024, use_cache = TRUE)
}
```

This vignette explores Colorado's public school enrollment data, surfacing key trends and demographic patterns. Colorado is part of the [state schooldata project](https://github.com/almartin82/njschooldata), a family of R packages providing consistent access to school enrollment data from all 50 states.

**Data years:** 2020 and 2024 (Student October Count, collected by the Colorado Department of Education). 881,000 students across 187 districts in the Centennial State.

---

## 1. Colorado lost 31,584 students since 2020

The pandemic accelerated enrollment decline in a state that had been growing for a decade. Colorado shed 3.5% of its student population between 2020 and 2024.

```{r statewide-trend}
state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

stopifnot(nrow(state_totals) > 0)
state_totals
```

```{r statewide-chart}
print(state_totals)

ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#003366") +
  geom_point(size = 3, color = "#003366") +
  geom_vline(xintercept = 2020.5, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 2020.5, y = max(state_totals$n_students) * 0.99,
           label = "COVID", hjust = -0.1, size = 3) +
  scale_y_continuous(labels = scales::comma, limits = c(850000, NA)) +
  labs(
    title = "Colorado Public School Enrollment (2020 vs 2024)",
    subtitle = "31,584 fewer students -- a 3.5% decline",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 2. Denver lost 3,877 students but remains the largest district

Denver County 1 dropped from 92,112 to 88,235 students, a 4.2% decline. Despite the losses, it is still Colorado's biggest district by a wide margin.

```{r denver-decline}
denver <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Denver County", district_name)) |>
  select(end_year, district_name, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

stopifnot(nrow(denver) > 0)
denver
```

```{r top-districts-chart}
top4 <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Denver County|Douglas County|Jefferson County|Cherry Creek", district_name))

stopifnot(nrow(top4) > 0)
print(top4 |> select(end_year, district_name, n_students))

ggplot(top4, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Colorado's Largest Districts: All Declining",
    subtitle = "Denver, Jeffco, Douglas County, and Cherry Creek all lost students",
    x = "School Year",
    y = "Enrollment",
    color = "District"
  )
```

---

## 3. Jefferson County lost 7,860 students -- shrinking faster than Denver

Jeffco shed 9.4% of its enrollment between 2020 and 2024, nearly 8,000 students. That is a steeper percentage decline than Denver.

```{r jeffco}
jeffco <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Jefferson County", district_name)) |>
  select(end_year, district_name, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

stopifnot(nrow(jeffco) > 0)
jeffco
```

```{r jeffco-chart}
front_range <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Jefferson County|Denver County|Douglas County", district_name))

print(front_range |> select(end_year, district_name, n_students))

ggplot(front_range, aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Front Range Giants: Enrollment Trends",
    subtitle = "Jeffco's 9.4% decline outpaces Denver's 4.2%",
    x = "School Year",
    y = "Enrollment",
    color = "District"
  )
```

---

## 4. Hispanic students are now 35.5% of enrollment

Hispanic students make up the second-largest group in Colorado, just behind white students. The Hispanic share grew from 33.9% in 2020 to 35.5% in 2024.

```{r demographics}
demographics <- enr_2024 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("hispanic", "white", "black", "asian",
                         "native_american", "multiracial", "pacific_islander")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(subgroup, n_students, pct) |>
  arrange(desc(n_students))

stopifnot(nrow(demographics) > 0)
demographics
```

```{r demographics-chart}
print(demographics)

demographics |>
  mutate(subgroup = forcats::fct_reorder(subgroup, n_students)) |>
  ggplot(aes(x = n_students, y = subgroup, fill = subgroup)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Colorado Student Demographics (2024)",
    subtitle = "Hispanic students are over a third of enrollment",
    x = "Number of Students",
    y = NULL
  )
```

---

## 5. White share dropped below 52.9% to 50.5% in four years

White students are still the largest group but their share fell 2.4 percentage points while the multiracial population surged from 4.5% to 5.3%.

```{r demographic-shift}
demo_shift <- enr |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("hispanic", "white", "black", "asian", "multiracial")) |>
  select(end_year, subgroup, n_students, pct) |>
  mutate(pct = round(pct * 100, 1))

stopifnot(nrow(demo_shift) > 0)
demo_shift
```

---

## 6. 261 charter schools serve 135,223 students

Colorado has one of the most expansive charter sectors in the country. Charter schools now enroll 15.3% of all public school students.

```{r charters}
charters <- enr_2024 |>
  filter(is_school, is_charter, subgroup == "total_enrollment", grade_level == "TOTAL")

state_total <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  pull(n_students)

charter_summary <- tibble(
  sector = c("All Public Schools", "Charter Schools"),
  enrollment = c(state_total, sum(charters$n_students, na.rm = TRUE)),
  pct = c(100, round(sum(charters$n_students, na.rm = TRUE) / state_total * 100, 1))
)

stopifnot(charter_summary$enrollment[2] > 0)
charter_summary
```

```{r charter-top}
top_charters <- charters |>
  arrange(desc(n_students)) |>
  head(5) |>
  select(district_name, campus_name, n_students)

print(top_charters)
```

---

## 7. Adams-Arapahoe (Aurora) is Colorado's most diverse district

Aurora Public Schools has no racial majority. Hispanic students make up 57.3%, followed by Black (16.8%), White (13.7%), Multiracial (5.9%), and Asian (4.8%).

```{r aurora}
aurora <- enr_2024 |>
  filter(is_district, grade_level == "TOTAL",
         grepl("Adams-Arapahoe", district_name),
         subgroup %in% c("hispanic", "white", "black", "asian", "multiracial")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(district_name, subgroup, n_students, pct) |>
  arrange(desc(pct))

stopifnot(nrow(aurora) > 0)
aurora
```

---

## 8. Nearly half of Colorado students qualify for free or reduced lunch

45.2% of Colorado students -- 398,112 children -- are economically disadvantaged, qualifying for free or reduced-price lunch.

```{r frl-state}
frl <- enr_2024 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup == "free_reduced_lunch") |>
  mutate(pct = round(pct * 100, 1)) |>
  select(subgroup, n_students, pct)

stopifnot(nrow(frl) > 0)
frl
```

```{r frl-districts}
frl_districts <- enr_2024 |>
  filter(is_district, subgroup == "free_reduced_lunch", grade_level == "TOTAL") |>
  mutate(pct = round(pct * 100, 1)) |>
  filter(n_students >= 1000) |>
  arrange(desc(pct)) |>
  head(10) |>
  select(district_name, n_students, pct)

print(frl_districts)

frl_districts |>
  mutate(district_name = forcats::fct_reorder(district_name, pct)) |>
  ggplot(aes(x = pct, y = district_name)) +
  geom_col(fill = "#E65100") +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Most Economically Disadvantaged Districts (2024)",
    subtitle = "Free/Reduced Lunch rate among districts with 1,000+ students",
    x = "FRL %",
    y = NULL
  )
```

---

## 9. Byers 32J grew 175% in four years

Byers 32J, northeast of Denver, surged from 2,344 to 6,456 students -- the fastest growth rate of any district in Colorado.

```{r biggest-gainers}
changes <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, district_name, n_students) |>
  pivot_wider(names_from = end_year, values_from = n_students, values_fn = sum) |>
  filter(!is.na(`2020`) & !is.na(`2024`)) |>
  mutate(change = `2024` - `2020`,
         pct_change = round(change / `2020` * 100, 1)) |>
  filter(`2020` >= 1000)

gainers <- changes |>
  arrange(desc(pct_change)) |>
  head(10)

stopifnot(nrow(gainers) > 0)
gainers
```

---

## 10. 9 of the 10 largest districts lost students

Academy 20 was the sole top-10 district to hold steady (gaining 4 students). Everyone else shrank, led by Jefferson County (-9.4%) and Adams 12 Five Star (-9.4%).

```{r top10-compare}
top10_names <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  pull(district_name)

top10_compare <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         district_name %in% top10_names) |>
  select(end_year, district_name, n_students) |>
  pivot_wider(names_from = end_year, values_from = n_students, values_fn = sum) |>
  mutate(change = `2024` - `2020`,
         pct_change = round(change / `2020` * 100, 1)) |>
  arrange(pct_change)

stopifnot(nrow(top10_compare) == 10)
top10_compare
```

```{r top10-chart}
print(top10_compare)

top10_compare |>
  mutate(district_name = forcats::fct_reorder(district_name, pct_change)) |>
  ggplot(aes(x = pct_change, y = district_name,
             fill = ifelse(pct_change >= 0, "gain", "loss"))) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct_change, "%")),
            hjust = ifelse(top10_compare$pct_change >= 0, -0.1, 1.1)) +
  scale_fill_manual(values = c("gain" = "#4CAF50", "loss" = "#F44336")) +
  labs(
    title = "Top 10 Districts: Change from 2020 to 2024",
    subtitle = "Only Academy 20 held steady; the rest all shrank",
    x = "% Change",
    y = NULL
  )
```

---

## 11. 115 tiny districts serve fewer than 1,000 students each

Colorado has an extraordinarily fragmented school system. Over 60% of districts enroll fewer than 1,000 students, though they educate a small fraction of the total.

```{r district-sizes}
size_cats <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(size = case_when(
    n_students >= 20000 ~ "Large (20K+)",
    n_students >= 5000 ~ "Medium (5K-20K)",
    n_students >= 1000 ~ "Small (1K-5K)",
    TRUE ~ "Tiny (<1K)"
  )) |>
  group_by(size) |>
  summarize(
    n_districts = n(),
    total_students = sum(n_students),
    .groups = "drop"
  ) |>
  mutate(pct_students = round(total_students / sum(total_students) * 100, 1))

stopifnot(nrow(size_cats) > 0)
size_cats
```

```{r smallest-districts}
smallest <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(n_students) |>
  head(10) |>
  select(district_name, n_students)

print(smallest)
```

---

## 12. Las Animas RE-1 lost 60% of its enrollment

The starkest decline in Colorado. Las Animas RE-1 went from 2,406 students in 2020 to just 956 in 2024 -- a loss of 1,450 students.

```{r biggest-losers}
losers <- changes |>
  arrange(pct_change) |>
  head(10)

stopifnot(nrow(losers) > 0)
losers
```

---

## 13. Adams County 14 has 86% economically disadvantaged students

Adams County 14, in the northern Denver metro area, has the highest free/reduced lunch rate among districts with 1,000+ students.

```{r frl-detail}
frl_top5 <- enr_2024 |>
  filter(is_district, subgroup == "free_reduced_lunch", grade_level == "TOTAL") |>
  mutate(pct = round(pct * 100, 1)) |>
  filter(n_students >= 1000) |>
  arrange(desc(pct)) |>
  head(5) |>
  select(district_name, n_students, pct)

stopifnot(nrow(frl_top5) > 0)
frl_top5
```

---

## 14. Colorado's gender split: 51.3% male, 48.7% female

Like most states, Colorado enrolls slightly more boys than girls -- a gap of about 23,000 students.

```{r gender-balance}
gender <- enr_2024 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("male", "female")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(subgroup, n_students, pct)

stopifnot(nrow(gender) == 2)
gender
```

```{r gender-chart}
print(gender)

gender |>
  mutate(subgroup = factor(subgroup, levels = c("female", "male"))) |>
  ggplot(aes(x = subgroup, y = n_students, fill = subgroup)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(scales::comma(n_students), "\n(", pct, "%)")),
            vjust = -0.2, size = 4) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = c("female" = "#E91E63", "male" = "#2196F3")) +
  labs(
    title = "Colorado Student Gender Distribution (2024)",
    subtitle = "23,381 more boys than girls enrolled",
    x = NULL,
    y = "Number of Students"
  )
```

---

## 15. Top 10 districts serve 55% of all students

Just 10 districts out of 187 educate more than half of Colorado's public school students, showing extreme concentration of enrollment in the Front Range metro areas.

```{r top-10-districts}
district_totals <- enr_2024 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, n_students)

state_total_val <- enr_2024 |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  pull(n_students)

top_10_pct <- round(sum(district_totals$n_students) / state_total_val * 100, 1)

result <- district_totals |>
  mutate(pct_of_state = round(n_students / state_total_val * 100, 1))

stopifnot(nrow(result) == 10)
result
```

```{r top-10-chart}
print(result)

district_totals |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#673AB7") +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Colorado's 10 Largest School Districts (2024)",
    subtitle = paste0("These districts serve ", top_10_pct, "% of all students"),
    x = "Total Enrollment",
    y = NULL
  )
```

---

## Summary

Colorado's school enrollment data reveals:

- **Shrinking enrollment**: 31,584 fewer students since 2020 (-3.5%)
- **Urban decline**: Denver, Jeffco, and all but one top-10 district losing students
- **Demographic shift**: White share fell to 50.5%, Hispanic grew to 35.5%
- **Charter expansion**: 261 charter schools enroll 15.3% of students
- **Economic disparity**: 45.2% of students qualify for free/reduced lunch
- **Fragmented system**: 115 tiny districts under 1,000 students
- **Concentrated enrollment**: Top 10 districts serve 55% of students

---

*Data sourced from the Colorado Department of Education Student October Count.*

---

## Session Info

```{r session-info}
sessionInfo()
```
